An iterated Grothendieck construction of semi-cubical types

### 2021-10-26 paper/paper.tex
[paper] Overture

% Lettered footnotes
\renewcommand{\thefootnote}{\alph{footnote}}

% Theorems
\newtheorem{notation}{Notation}
\newtheorem{definition}{Definition}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{remark}{Remark}
\newtheorem{example}{Example}

@@

% Roman numerals for chapters
\renewcommand{\thesection}{\Roman{section}}

@@

% Fonts
\setmainfont{Avenir Next}
\setmonofont{Source Code Pro}
\newfontfamily\mainfont{Avenir Next}
\newfontfamily\caslon{Adobe Caslon Pro}
\defaultfontfeatures{Ligatures=TeX, Scale=MatchUppercase}

@@

\section{Overture}
Cubical type theory is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. Cubical type theory has several attractive features, and people have been working on it since 2014~\cite{Bezem14}. Today, an experimental branch of Agda, known as \emph{Cubical Agda} is being worked on~\cite{Vezzosi21}.

Our contribution uses the indexed approach to constructing semi-cubical types, as opposed to CCHM~\cite{Cohen16} in 2016, which takes a fibered approach. We will expound on this further in the Grothendieck construction section. We also use \emph{unicity of identity of proofs} (UIP) to finish off the last coherence condition, and we have dedicated a section to discussing it in various settings.

@@

\begin{definition}[$\CSet$ in terms of objects and morphisms]
  \begin{align*}
    \obj(\CSet) & := X_n                                                   \\
    \mor(\CSet) & := X_\lambda, \text{where \lambda is \Cube{}-morphism}
  \end{align*}

@@

\section{Grothendieck construction}
The Grothendieck construction provides a correspondence between a fibered representation, and an indexed representation.

@@

\section{Unicity of identity proofs}
We dedicate this section to discussing UIP in different settings, and supply intuitions into this notion.

@@

\begin{listing}[H]
  \begin{minted}[mathescape]{coq}
    Inductive SET : Type :=
      node : \forall A : Type, (A -> SET) -> SET
      ~$=_\textrm{SET}$~ : ...
  \end{minted}
\end{listing}

[draw tree structure of set]

@@

In attempting to write the proof without UIP, we'd have a relation between \mintinline{coq}{cohNbox} and \mintinline{coq}{cohPredNbox}, which would be very difficult to prove.

@@

>>>>

@@

% Use lettered footnotes
\renewcommand{\thefootnote}{\alph{footnote}}

% amsthm theorems
\swapnumbers
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{terminology}[definition]{Terminology}
\newtheorem{notation}[definition]{Notation}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{remark}[definition]{Remark}
\newtheorem{example}[definition]{Example}
\newtheorem{caution}[definition]{Caution}
\newtheorem{discussion}[definition]{Discussion}

\usepackage{thmtools}

@@

% Fonts
\setmainfont{Avenir Next}
\setmonofont{Source Code Pro}
\newfontfamily\mainfont{Avenir Next}
\defaultfontfeatures{Ligatures=TeX, Scale=MatchUppercase}

@@

\section{Overture}
Cubical type theory~\cite{Bezem14}~\cite{Cohen16}~\cite{Angiuli17} is an extension of dependent type theory that provides a computational interpretation of Voevodsky's \emph{univalence axiom}, in a field known as \emph{homotopy type theory}, or HoTT. Two real-world projects that implement cubical type theory are worthy of mention: an experimental branch of Agda, known as \emph{Cubical Agda}~\cite{Vezzosi21}, and the Arend theorem prover~\footnote{\href{https://arend-lang.github.io}{arend-lang.github.io}}. Cubical type theory is usually modeled as cubical sets, in the set-theoretic setting. Efforts to model cubical type theory have been ongoing since 2014, and many of these efforts are in the set-theoretic setting.

Our contribution is to model the core of cubical types, semi-cubical sets, in type theory, using an indexed representation, in contrast to previous efforts that use a fibered representation. We use an \emph{iterated reverse-Grothendieck} construction as the basis of our formalization, which we expound on in \ref{sec:grothendieckconstr}. Another feature of our work is that, as we are in the type-theoretic setting, we use \emph{unicity of identity of proofs} (UIP), which would be an inherent property in the set-theoretic setting (see \ref{sec:uip}).

@@

\begin{definition}[$\CSet$ in terms of objects and morphisms]
  \begin{align*}
    \obj(\CSet) & := X_n                                                   \\
    \mor(\CSet) & := X_\lambda, \text{where $\lambda$ is \Cube{}-morphism}
  \end{align*}

@@

\section{Grothendieck construction\label{sec:grothendieckconstr}}
The Grothendieck construction provides a correspondence between a fibered representation, and an indexed representation.

@@

\section{Unicity of identity proofs\label{sec:uip}}
We dedicate this section to discussing UIP in different settings, and supply intuitions into this notion.

@@

\begin{listing}[H]
  \begin{minted}[mathescape]{coq}
    Inductive SET : Type :=
      node : ~$\forall$~A : Type, (A -> SET) -> SET
      ~$=_\textrm{SET}$~ : ...
  \end{minted}
\end{listing}

The special case of \mintinline{coq}{(Bool -> SET) -> SET} can be drawn as:

\begin{equation*}
  \begin{tikzcd}
    & & \{0, 1, 2\} \arrow[dr]\arrow[dl] & \\
    & \{0, 1\} \arrow[dl]\arrow[dr] & & \{2\} \\
    \{0\} & & \{1\} & \\
  \end{tikzcd}
\end{equation*}

In general, \mintinline{coq}{SET} can have an $n$-ary branching structure, splitting out the various subsets until we get to the leaves.

@@

In attempting to write the proof without UIP, we'd have a relation between \mintinline{coq}{cohNbox} and \mintinline{coq}{cohPredNbox}, which, we expect to be tricky to prove.

@@

  \bibitem[Angiuli]{Angiuli17}
  Carlo Angiuli, Guillaume Brunerie, Thierry Coquand, Kuen-Bang Hou (Favonia),Robert Harper, \& Daniel R. Licata. arXiv preprint.
